---
title: "P-values two ways"
---

```{r}
#| label: setup
#| message: false
#| echo: false

library(tidyverse)
set.seed(222)
theme_set(theme_bw())

# Read the soil core data
soil_cores <- read_csv('/courses/EDS222/data/exide-soil-cores.csv')
```

## Permutation test

In the lab, you used a **permutation test** to investigate the claim _"Lead is elevated in the drip line compared to the rest of the yard"_. In this activity, you will investigate the claim _"Soil samples with lead levels exceeding 200ppm are more prevalent north of Exide than to the south"_. Use the _difference in proportions_ as the statistic to test this claim.

### State your hypotheses

State the null and alternative hypotheses for this claim.

$H_0$: There is no difference in the occurence of lead levels exceeding 200 ppm between samples from the north and samples from the south

$H_A$: Soil samples with lead levels exceeding 200ppm are more prevalent north of Exide than to the south  

### Calculate the observed value of the statistic

Fill in the code chunk below to calculate the observed value of the statistic

::: {.callout-tip}

Use `dplyr` functions to wrangle the soil core data for this part. Properties starting with "ENA" are in the extended north area, and the "ESA" prefix means extended south area. `str_sub()` will help here.

:::

```{r}
ena_esa <- soil_cores |>
  mutate(area = str_sub(property_id, start = 1L, end = 3L)) |>  # create a column called "area", which is the first three characters of property_id
  filter((area == 'ENA') | (area == 'ESA'))  # then filter to observations where area is ENA or ESA

# summarize to get the proportion of samples where lead is over 200 ppm, by area, sort by the area
area_summ <- ena_esa |> summarize(prop200 = mean(pb_ppm > 200), .by = area) |>
  arrange(area)

obs_diff_props <- area_summ$prop200[1] - area_summ$prop200[2] # get the difference in proportions between the two groups

obs_diff_props
```

### Shuffle the predictor variable and recalculate the statistic

Shuffle the predictor variable and recalculate the statistic. Do it just once - we'll repeat in the next section.

```{r}
# summarize to get the proportion of samples where lead is over 200 ppm, by area, sort by the area
shuffled_summ <- ena_esa |>
  mutate(area = sample(area)) |> # shuffle the predictor (use mutate and sample)
  summarize(prop200 = mean(pb_ppm > 200), .by = area) |> 
  arrange(area)

shuffled_diff_props <- shuffled_summ$prop200[1] - shuffled_summ$prop200[2] # get the difference in proportions between the two groups

shuffled_diff_props
```

### Repeat shuffling

Repeat shuffling 10,000 times to create a vector that simulates the null distribution.

```{r}
area_null_dist <- replicate(
  1e4,
  {
  shuffled_summ <- ena_esa |>
  mutate(area = sample(area)) |> 
  summarize(prop200 = mean(pb_ppm > 200), .by = area) |> 
  arrange(area)

  shuffled_summ$prop200[1] - shuffled_summ$prop200[2]
  }
)
```

### Compare the observed to the null

Visualize the null distribution using a histogram and add a vertical line indicating the observed statistic. Indicate which part of the null distribution is greater than or equal to the observed statistic using fill.

```{r}
tibble(diff_props = area_null_dist) |>
  ggplot(aes(diff_props)) +
  geom_histogram(binwidth = .01) + 
  geom_vline(xintercept = obs_diff_props, color = "magenta", linewidth = 1.5)

  # set the aesthetics for x and fill using ggplot
  # add a histogram
  # add a vertical line at the observed difference in proportions
```

Calculate your p-value.

```{r}
pval <- mean(area_null_dist > obs_diff_props) # the proportion of null distribution values greater than or equal to the observed statistic

pval
```

## Mathematical model

Now, let's repeat the process to estimate the p-value, but use the mathematical model instead.

The steps **State your hypotheses** and **Calculate the observed value of the statistic** are unchanged.

### Calculate the standard error

The standard error of the difference in proportions for a null distribution is:

$$
SE = \sqrt{\hat p_{pool} (1 - \hat p_{pool}) \left( \frac{1}{n_1} + \frac{1}{n_2} \right) }
$$

Fill in the code chunk below to calculate the standard error.

```{r}
p_pool <- mean(ena_esa$pb_ppm > 200) # total "successes" (soil lead over 200) over total observations
n1 <- # number of observations in the "north" group
n2 <- # number of observations in the "south" group
se <- # use the formula above
```

### Compare the observed to the null

Estimate the null distribution as a normal distribution, with a mean of 0 and a standard deviation equal to the standard error. 

Visualize the null distribution and the observed value of the null statistic.

```{r}
# Create a tibble for the null distribution
null_dist <- tibble(
  # create a column called "diff_props", a sequence from -3SE to 3SE with length.out of 1000
  # then create a column called "null_pdf", the PDF of the normal distribution with the right mean and standard deviation
)

# Visualize the null distribution using a line geometry, then add a vertical line indicating the observed value of 
# the statistic
```

Calculate your p-value. Once again, it's the area under the curve. But instead of counting up permutations, you need to integrate the PDF. Don't worry - you don't have to do it by hand! `pnorm()` is the area under the curve for the normal distribution.

Here's an example of how it works. Given a normal distribution with mean 0 and sd 2, the area under the curve _to the left_ of -2 is:

```{r}
pnorm(-2, mean = 0, sd = 2)
```

By default, `pnorm()` gets the area under the curve to the left of a value. To get the area under the curve _to the right_ of a value, set the parameter `lower.tail` to `FALSE`.

```{r}
pnorm(-2, mean = 0, sd = 2, lower.tail = FALSE)
```

Now, fill in the code chunk below to calculate your p-value.

```{r}
pval <- pnorm(your_observed_statistic, mean = ???, sd = ???, lower.tail = ???)
```

## P-values two ways

How do your two p-values compare?
